% !TeX root = owstl.tex

\chapter{Type Traits}

\section{Introduction}

The header \filepath{type_traits} is based on the metaprogramming section of
n1836 ``Draft Technical Report on C++ Library Extensions.'' It contains a set
of templates that allow compile-time testing and modification of types.

\section{Status}

About half of the required functionality has been implemented so far. There
are currently a few compiler bugs stopping some parts being implemented.

The missing templates are:

\begin{enumerate}
\item \code{is_member_object_pointer}
\item \code{is_member_function_pointer}
\item \code{is_enum}
\item \code{is_union}
\item \code{is_class}
\item \code{is_function}
\item \code{struct is_object}
\item \code{is_scalar}
\item \code{is_compound}
\item \code{is_member_pointer}
\item \code{is_pod}
\item \code{is_empty}
\item \code{is_polymorphic}
\item \code{is_abstract}
\item \code{has_trivial_constructor}
\item \code{has_trivial_copy}
\item \code{has_trivial_assign}
\item \code{has_trivial_destructor}
\item \code{has_nothrow_constructor}
\item \code{has_nothrow_copy}
\item \code{has_nothrow_assign}
\item \code{has_virtual_destructor}
\item \code{is_signed}
\item \code{is_unsigned}
\item \code{alignment_of}
\item \code{rank}
\item \code{extent}
\item \code{is_same}
\item \code{is_base_of}
\item \code{is_convertible}
\item \code{remove_extent}
\item \code{remove_all_extents}
\item \code{add_pointer}
\item \code{aligned_storage}
\end{enumerate}

\section{Design Details}

\subsection{Querying Types}

This is implemented by specializing templates for the types that the test
holds true. The class derives from a helper class that contains a static const
value. The important cases are when this static const is a bool and is true or
false. The user can then access \code{is_void<type>::value} to see if the test
is positive. A set of macros are used to help make the definitions look a bit
less cluttered. There is a default case which declares the main template and
is usually false. There are then other macros that define the specializations.
There are also macros that define 4 specializations for the const volatile
qualified variations of the type.

\subsection{Modifying Types}

This works similarly. The template is specialized for the type with the
modifier and the class contains a typedef type that refers to the modified
type. Macros aren't used for the modifiers as they tend to have subtle
differences for each template and in many cases there isn't the need for 4
different CV variations.

\subsection{Use in Main Library}

This header should be a help for writing the constructors and member functions
of the standard containers that are required to have different behavior for
iterators and integral types.
